---
title: "DATA 624 Project 1"
subtitle: "Exercise 1, Exercise 2, Bonus #1, and Bonus #2"
author: "Group 3"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  word_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5
)

set.seed(624)

required_packages <- c(
  "forecast",
  "ggplot2",
  "dplyr",
  "tidyr",
  "knitr",
  "lubridate"
)

missing_packages <- required_packages[!required_packages %in% rownames(installed.packages())]

if (length(missing_packages) > 0) {
  stop(
    paste0(
      "Please install these packages first: ",
      paste(missing_packages, collapse = ", "),
      "\nRun: install.packages(c('",
      paste(missing_packages, collapse = "', '"),
      "'))"
    )
  )
}

library(forecast)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(lubridate)
```

# Executive Summary

This project uses time series forecasting to support planning decisions for two organizations.

For **Lone Star Appliances**, the main question is how many units the company should expect to ship in the next several quarters. This matters for production, warehouse inventory, staffing, and logistics contracts.

For **Texas HealthLink**, the main question is how many patient visits each clinic should expect over the next 12 months. This matters for staffing, budgeting, scheduling, and facility planning.

Both datasets show growth over time and seasonal movement. Both also show disruption around the COVID period. Because of that, I would use the forecasts as planning baselines, not as exact promises. The best use of this report is to help leadership plan capacity, prepare for busy periods, and understand where the risk is.

# Exercise #1: Lone Star Appliances

## Problem and Approach

For Lone Star, the forecast is used as a planning tool for a manufacturer. The main point is not just the model output. The main point is what the forecast means for production, staffing, inventory, and logistics.

The Board should be able to answer three questions after reading this section:

1. What is the shipment pattern?
2. What should we expect next?
3. What should the company do with the forecast?


## Board-Level Summary

Lone Star Appliances has a clear upward shipment pattern. Shipments generally increased from 2010 through 2021. The company also has a seasonal pattern. The fourth quarter is usually the strongest quarter, which makes sense for a business that ships appliances to retailers.

The main issue is the 2020 disruption. Shipments dropped in 2020 Q2, then recovered. That tells me the business has a strong base demand, but it is still exposed to outside shocks. A forecast can help the Board plan, but it cannot remove risk.

My recommendation is to use the final forecast as the baseline production plan. I would not plan staffing, inventory, or carrier contracts only around the average forecast. I would also keep a buffer for Q4 and for any quarter where demand moves above the expected range.

## Business Decisions Supported

The Lone Star forecast supports these decisions:

- production planning for the next 1 to 3 years
- staffing needs by quarter
- warehouse inventory targets
- logistics contracts with carriers
- risk planning for high-demand quarters

## Main Risk

The biggest risk is that future demand may not follow the same pattern as the past. COVID showed that one shock can change shipments quickly. Supply chain issues, retailer orders, labor constraints, or economic changes could also change the forecast.

## Exercise #1 Technical Work

```{r lone-star-load-data}
lone_star_text <- "year,quarter,shipments
2010,1,60
2010,2,63
2010,3,66
2010,4,70
2011,1,62
2011,2,65
2011,3,69
2011,4,73
2012,1,65
2012,2,69
2012,3,73
2012,4,78
2013,1,68
2013,2,72
2013,3,77
2013,4,82
2014,1,72
2014,2,77
2014,3,82
2014,4,88
2015,1,82
2015,2,88
2015,3,93
2015,4,101
2016,1,86
2016,2,92
2016,3,98
2016,4,107
2017,1,90
2017,2,97
2017,3,104
2017,4,114
2018,1,95
2018,2,103
2018,3,111
2018,4,122
2019,1,100
2019,2,108
2019,3,117
2019,4,129
2020,1,96
2020,2,85
2020,3,110
2020,4,124
2021,1,107
2021,2,116
2021,3,126
2021,4,139"

lone_star <- read.csv(text = lone_star_text)

lone_star$quarter_label <- paste0(lone_star$year, " Q", lone_star$quarter)
lone_star$time_index <- seq_len(nrow(lone_star))
lone_star$quarter_factor <- factor(lone_star$quarter)

head(lone_star)
```

```{r lone-star-summary}
summary(lone_star$shipments)
```

```{r lone-star-plot}
library(ggplot2)

p_lone_star <- ggplot(lone_star, aes(x = time_index, y = shipments)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(
    breaks = seq(1, nrow(lone_star), by = 4),
    labels = lone_star$year[seq(1, nrow(lone_star), by = 4)]
  ) +
  labs(
    title = "Lone Star Appliances Quarterly Shipments",
    x = "Year",
    y = "Shipments (thousands of units)"
  )

p_lone_star
```

```{r lone-star-ts}
library(forecast)

lone_star_ts <- ts(lone_star$shipments, start = c(2010, 1), frequency = 4)

plot(lone_star_ts,
     main = "Lone Star Quarterly Shipments",
     ylab = "Shipments (thousands)",
     xlab = "Year")
```

```{r lone-star-decomposition}
lone_star_decomp <- stl(lone_star_ts, s.window = "periodic")
plot(lone_star_decomp)
```

This decomposition separates the long-term trend, the quarterly seasonal movement, and the remaining noise. That helps the Board see that the forecast is not just one line. It is based on a pattern that repeats, but still has risk.

```{r lone-star-acf-pacf}
Acf(lone_star_ts, main = "ACF: Lone Star Quarterly Shipments")
Pacf(lone_star_ts, main = "PACF: Lone Star Quarterly Shipments")
```

The ACF and PACF plots are used as a technical check. They help show whether there is leftover time-series structure that a model such as ARIMA should consider.

```{r lone-star-models}
set.seed(624)

ls_snaive <- snaive(lone_star_ts, h = 12)
ls_ets <- ets(lone_star_ts)
ls_ets_fc <- forecast(ls_ets, h = 12)
ls_arima <- auto.arima(lone_star_ts, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
ls_arima_fc <- forecast(ls_arima, h = 12)

ls_reg <- tslm(lone_star_ts ~ trend + season)
ls_reg_fc <- forecast(ls_reg, h = 12)

ls_models <- list(
  Seasonal_Naive = ls_snaive,
  ETS = ls_ets_fc,
  ARIMA = ls_arima_fc,
  Regression = ls_reg_fc
)

ls_ets
ls_arima
summary(ls_reg)
```

```{r lone-star-accuracy}
ls_accuracy <- rbind(
  Seasonal_Naive = accuracy(ls_snaive)[1, c("RMSE", "MAE", "MAPE")],
  ETS = accuracy(ls_ets_fc)[1, c("RMSE", "MAE", "MAPE")],
  ARIMA = accuracy(ls_arima_fc)[1, c("RMSE", "MAE", "MAPE")],
  Regression = accuracy(ls_reg_fc)[1, c("RMSE", "MAE", "MAPE")]
)

knitr::kable(round(ls_accuracy, 3), caption = "Lone Star in-sample model accuracy comparison")
```

The table above is useful, but it is not the only model check. I also used a holdout test by training models through 2020 Q4 and testing them on the 2021 quarters. This is closer to how the forecast would be used in practice.

```{r lone-star-holdout-validation}
ls_train <- window(lone_star_ts, end = c(2020, 4))
ls_test <- window(lone_star_ts, start = c(2021, 1))
ls_h <- length(ls_test)

ls_holdout_snaive <- snaive(ls_train, h = ls_h)
ls_holdout_ets <- forecast(ets(ls_train), h = ls_h)
ls_holdout_arima <- forecast(auto.arima(ls_train, seasonal = TRUE), h = ls_h)
ls_holdout_reg <- forecast(tslm(ls_train ~ trend + season), h = ls_h)

ls_holdout_accuracy <- rbind(
  Seasonal_Naive = accuracy(ls_holdout_snaive, ls_test)["Test set", c("RMSE", "MAE", "MAPE")],
  ETS = accuracy(ls_holdout_ets, ls_test)["Test set", c("RMSE", "MAE", "MAPE")],
  ARIMA = accuracy(ls_holdout_arima, ls_test)["Test set", c("RMSE", "MAE", "MAPE")],
  Regression = accuracy(ls_holdout_reg, ls_test)["Test set", c("RMSE", "MAE", "MAPE")]
)

knitr::kable(round(ls_holdout_accuracy, 3), caption = "Lone Star holdout validation using 2021 as the test year")
```

The holdout test is important because it checks the models on data that was not used to fit them. I used RMSE as the main metric because it keeps larger misses visible.

```{r lone-star-select-champion}
ls_accuracy_df <- data.frame(
  model = rownames(ls_holdout_accuracy),
  RMSE = ls_holdout_accuracy[, "RMSE"],
  MAE = ls_holdout_accuracy[, "MAE"],
  MAPE = ls_holdout_accuracy[, "MAPE"],
  row.names = NULL
)

ls_champion_name <- ls_accuracy_df$model[which.min(ls_accuracy_df$RMSE)]
ls_champion_name
```

```{r lone-star-forecast-table}
ls_forecast <- if (ls_champion_name == "ETS") {
  ls_ets_fc
} else if (ls_champion_name == "ARIMA") {
  ls_arima_fc
} else if (ls_champion_name == "Regression") {
  ls_reg_fc
} else {
  ls_snaive
}

ls_fc_table <- data.frame(
  period = paste0(rep(2022:2024, each = 4), " Q", rep(1:4, times = 3)),
  forecast = as.numeric(ls_forecast$mean),
  lower_80 = as.numeric(ls_forecast$lower[, 1]),
  upper_80 = as.numeric(ls_forecast$upper[, 1]),
  lower_95 = as.numeric(ls_forecast$lower[, 2]),
  upper_95 = as.numeric(ls_forecast$upper[, 2])
)

ls_fc_table_print <- ls_fc_table
ls_fc_table_print[, c("forecast", "lower_80", "upper_80", "lower_95", "upper_95")] <-
  round(ls_fc_table_print[, c("forecast", "lower_80", "upper_80", "lower_95", "upper_95")], 1)

knitr::kable(ls_fc_table_print, caption = "Lone Star 12-quarter forecast")
```

```{r lone-star-forecast-plot}
autoplot(ls_forecast) +
  labs(
    title = paste("Lone Star Forecast - Champion Model:", ls_champion_name),
    x = "Year",
    y = "Shipments (thousands)"
  )
```

```{r lone-star-residuals}
checkresiduals(if (ls_champion_name == "ETS") ls_ets else if (ls_champion_name == "ARIMA") ls_arima else ls_reg)
```

## Exercise #1 Recommendation

Lone Star should use the forecast as a baseline for production and logistics planning. The company should plan extra capacity around the fourth quarter because the data shows stronger demand near year-end. The company should also keep a reserve plan because COVID showed that one quarter can move far away from the normal pattern.

The Board should use the forecast as a planning baseline. Based on the historical pattern, demand is expected to keep rising with higher pressure in later quarters. I would not recommend planning only to the point forecast. I would use the forecast range because the company has already seen that shocks can change demand quickly.

# Exercise #2: Texas HealthLink

## Problem and Approach

For Texas HealthLink, the forecast is used as a staffing and capacity planning tool. This section is more detailed than Exercise #1 because the data has five clinics and a systemwide total.

The Board should be able to answer four questions after reading this section:

1. Which clinics have the highest expected demand?
2. What is the 12-month outlook for each clinic?
3. What is the systemwide patient volume outlook?
4. What staffing or facility risks should leadership watch?


## Board-Level Summary

Texas HealthLink has five clinics with monthly visit data from 2016 through 2023. The main business need is to forecast patient visits for each clinic and for the total system.

This forecast supports staffing, budget planning, facility planning, and scheduling. Since patient demand can change by clinic, the systemwide forecast alone is not enough. Leadership needs both the total forecast and the clinic-level view.

The early review shows that patient visits generally grow over time. Some clinics are larger than others, so staffing plans should not be spread evenly across all clinics. The COVID period also created volatility, which should be treated as a planning risk.

## Business Decisions Supported

The HealthLink forecast supports these decisions:

- monthly staffing plans
- clinic scheduling
- budget planning
- facility expansion decisions
- systemwide resource allocation
- risk planning for peak demand months

## Exercise #2 Technical Work

```{r healthlink-load-data}
healthlink_text <- "row,clinic,month,visits
1,Temple,2016-01-01,2059
2,Temple,2016-02-01,2028
3,Temple,2016-03-01,1930
4,Temple,2016-04-01,1770
5,Temple,2016-05-01,1797
6,Temple,2016-06-01,1943
7,Temple,2016-07-01,1921
8,Temple,2016-08-01,1923
9,Temple,2016-09-01,1815
10,Temple,2016-10-01,1877
11,Temple,2016-11-01,1841
12,Temple,2016-12-01,1796
13,Temple,2017-01-01,2001
14,Temple,2017-02-01,2087
15,Temple,2017-03-01,2005
16,Temple,2017-04-01,1964
17,Temple,2017-05-01,1839
18,Temple,2017-06-01,2021
19,Temple,2017-07-01,2063
20,Temple,2017-08-01,2058
21,Temple,2017-09-01,1974
22,Temple,2017-10-01,1938
23,Temple,2017-11-01,1933
24,Temple,2017-12-01,1931
25,Temple,2018-01-01,2127
26,Temple,2018-02-01,2071
27,Temple,2018-03-01,2107
28,Temple,2018-04-01,2009
29,Temple,2018-05-01,2143
30,Temple,2018-06-01,2081
31,Temple,2018-07-01,2142
32,Temple,2018-08-01,2141
33,Temple,2018-09-01,2006
34,Temple,2018-10-01,2068
35,Temple,2018-11-01,2167
36,Temple,2018-12-01,2115
37,Temple,2019-01-01,2218
38,Temple,2019-02-01,2199
39,Temple,2019-03-01,2109
40,Temple,2019-04-01,2188
41,Temple,2019-05-01,2040
42,Temple,2019-06-01,2260
43,Temple,2019-07-01,2339
44,Temple,2019-08-01,2145
45,Temple,2019-09-01,2202
46,Temple,2019-10-01,2152
47,Temple,2019-11-01,2082
48,Temple,2019-12-01,2093
49,Temple,2020-01-01,2216
50,Temple,2020-02-01,2288
51,Temple,2020-03-01,2240
52,Temple,2020-04-01,2007
53,Temple,2020-05-01,2100
54,Temple,2020-06-01,1985
55,Temple,2020-07-01,2367
56,Temple,2020-08-01,2374
57,Temple,2020-09-01,2276
58,Temple,2020-10-01,2203
59,Temple,2020-11-01,2265
60,Temple,2020-12-01,2263
61,Temple,2021-01-01,2442
62,Temple,2021-02-01,2394
63,Temple,2021-03-01,2483
64,Temple,2021-04-01,2290
65,Temple,2021-05-01,2383
66,Temple,2021-06-01,2345
67,Temple,2021-07-01,2340
68,Temple,2021-08-01,2618
69,Temple,2021-09-01,2327
70,Temple,2021-10-01,2378
71,Temple,2021-11-01,2406
72,Temple,2021-12-01,2347
73,Temple,2022-01-01,2535
74,Temple,2022-02-01,2534
75,Temple,2022-03-01,2507
76,Temple,2022-04-01,2412
77,Temple,2022-05-01,2414
78,Temple,2022-06-01,2632
79,Temple,2022-07-01,2468
80,Temple,2022-08-01,2454
81,Temple,2022-09-01,2450
82,Temple,2022-10-01,2475
83,Temple,2022-11-01,2490
84,Temple,2022-12-01,2444
85,Temple,2023-01-01,2536
86,Temple,2023-02-01,2684
87,Temple,2023-03-01,2595
88,Temple,2023-04-01,2452
89,Temple,2023-05-01,2498
90,Temple,2023-06-01,2588
91,Temple,2023-07-01,2675
92,Temple,2023-08-01,2621
93,Temple,2023-09-01,2589
94,Temple,2023-10-01,2522
95,Temple,2023-11-01,2573
96,Temple,2023-12-01,2549
97,Belton,2016-01-01,978
98,Belton,2016-02-01,942
99,Belton,2016-03-01,930
100,Belton,2016-04-01,996
101,Belton,2016-05-01,944
102,Belton,2016-06-01,914
103,Belton,2016-07-01,944
104,Belton,2016-08-01,925
105,Belton,2016-09-01,1024
106,Belton,2016-10-01,992
107,Belton,2016-11-01,933
108,Belton,2016-12-01,970
109,Belton,2017-01-01,1005
110,Belton,2017-02-01,1007
111,Belton,2017-03-01,998
112,Belton,2017-04-01,940
113,Belton,2017-05-01,1034
114,Belton,2017-06-01,1010
115,Belton,2017-07-01,1021
116,Belton,2017-08-01,1030
117,Belton,2017-09-01,1033
118,Belton,2017-10-01,967
119,Belton,2017-11-01,952
120,Belton,2017-12-01,1063
121,Belton,2018-01-01,1052
122,Belton,2018-02-01,1045
123,Belton,2018-03-01,1029
124,Belton,2018-04-01,961
125,Belton,2018-05-01,993
126,Belton,2018-06-01,1085
127,Belton,2018-07-01,1108
128,Belton,2018-08-01,1096
129,Belton,2018-09-01,1017
130,Belton,2018-10-01,1038
131,Belton,2018-11-01,1012
132,Belton,2018-12-01,1015
133,Belton,2019-01-01,1153
134,Belton,2019-02-01,1081
135,Belton,2019-03-01,1204
136,Belton,2019-04-01,1056
137,Belton,2019-05-01,1073
138,Belton,2019-06-01,1078
139,Belton,2019-07-01,1089
140,Belton,2019-08-01,1063
141,Belton,2019-09-01,1073
142,Belton,2019-10-01,1101
143,Belton,2019-11-01,1101
144,Belton,2019-12-01,1061
145,Belton,2020-01-01,1134
146,Belton,2020-02-01,1150
147,Belton,2020-03-01,1234
148,Belton,2020-04-01,913
149,Belton,2020-05-01,955
150,Belton,2020-06-01,1082
151,Belton,2020-07-01,1156
152,Belton,2020-08-01,1110
153,Belton,2020-09-01,1101
154,Belton,2020-10-01,1151
155,Belton,2020-11-01,1121
156,Belton,2020-12-01,1118
157,Belton,2021-01-01,1200
158,Belton,2021-02-01,1222
159,Belton,2021-03-01,1286
160,Belton,2021-04-01,1152
161,Belton,2021-05-01,1203
162,Belton,2021-06-01,1229
163,Belton,2021-07-01,1203
164,Belton,2021-08-01,1151
165,Belton,2021-09-01,1155
166,Belton,2021-10-01,1160
167,Belton,2021-11-01,1186
168,Belton,2021-12-01,1240
169,Belton,2022-01-01,1354
170,Belton,2022-02-01,1328
171,Belton,2022-03-01,1265
172,Belton,2022-04-01,1134
173,Belton,2022-05-01,1192
174,Belton,2022-06-01,1256
175,Belton,2022-07-01,1290
176,Belton,2022-08-01,1299
177,Belton,2022-09-01,1251
178,Belton,2022-10-01,1172
179,Belton,2022-11-01,1266
180,Belton,2022-12-01,1218
181,Belton,2023-01-01,1317
182,Belton,2023-02-01,1317
183,Belton,2023-03-01,1335
184,Belton,2023-04-01,1253
185,Belton,2023-05-01,1189
186,Belton,2023-06-01,1329
187,Belton,2023-07-01,1319
188,Belton,2023-08-01,1297
189,Belton,2023-09-01,1277
190,Belton,2023-10-01,1281
191,Belton,2023-11-01,1289
192,Belton,2023-12-01,1350
193,Killeen,2016-01-01,1385
194,Killeen,2016-02-01,1410
195,Killeen,2016-03-01,1466
196,Killeen,2016-04-01,1377
197,Killeen,2016-05-01,1387
198,Killeen,2016-06-01,1367
199,Killeen,2016-07-01,1502
200,Killeen,2016-08-01,1411
201,Killeen,2016-09-01,1447
202,Killeen,2016-10-01,1292
203,Killeen,2016-11-01,1367
204,Killeen,2016-12-01,1434
205,Killeen,2017-01-01,1432
206,Killeen,2017-02-01,1436
207,Killeen,2017-03-01,1433
208,Killeen,2017-04-01,1343
209,Killeen,2017-05-01,1380
210,Killeen,2017-06-01,1485
211,Killeen,2017-07-01,1373
212,Killeen,2017-08-01,1491
213,Killeen,2017-09-01,1488
214,Killeen,2017-10-01,1534
215,Killeen,2017-11-01,1503
216,Killeen,2017-12-01,1482
217,Killeen,2018-01-01,1454
218,Killeen,2018-02-01,1516
219,Killeen,2018-03-01,1534
220,Killeen,2018-04-01,1503
221,Killeen,2018-05-01,1469
222,Killeen,2018-06-01,1477
223,Killeen,2018-07-01,1630
224,Killeen,2018-08-01,1598
225,Killeen,2018-09-01,1510
226,Killeen,2018-10-01,1565
227,Killeen,2018-11-01,1443
228,Killeen,2018-12-01,1549
229,Killeen,2019-01-01,1586
230,Killeen,2019-02-01,1652
231,Killeen,2019-03-01,1621
232,Killeen,2019-04-01,1572
233,Killeen,2019-05-01,1613
234,Killeen,2019-06-01,1612
235,Killeen,2019-07-01,1669
236,Killeen,2019-08-01,1565
237,Killeen,2019-09-01,1534
238,Killeen,2019-10-01,1652
239,Killeen,2019-11-01,1601
240,Killeen,2019-12-01,1485
241,Killeen,2020-01-01,1616
242,Killeen,2020-02-01,1680
243,Killeen,2020-03-01,1739
244,Killeen,2020-04-01,1407
245,Killeen,2020-05-01,1449
246,Killeen,2020-06-01,1532
247,Killeen,2020-07-01,1774
248,Killeen,2020-08-01,1699
249,Killeen,2020-09-01,1639
250,Killeen,2020-10-01,1560
251,Killeen,2020-11-01,1659
252,Killeen,2020-12-01,1631
253,Killeen,2021-01-01,1707
254,Killeen,2021-02-01,1753
255,Killeen,2021-03-01,1819
256,Killeen,2021-04-01,1584
257,Killeen,2021-05-01,1669
258,Killeen,2021-06-01,1762
259,Killeen,2021-07-01,1717
260,Killeen,2021-08-01,1785
261,Killeen,2021-09-01,1735
262,Killeen,2021-10-01,1718
263,Killeen,2021-11-01,1603
264,Killeen,2021-12-01,1861
265,Killeen,2022-01-01,1818
266,Killeen,2022-02-01,1867
267,Killeen,2022-03-01,1854
268,Killeen,2022-04-01,1807
269,Killeen,2022-05-01,1803
270,Killeen,2022-06-01,1818
271,Killeen,2022-07-01,1853
272,Killeen,2022-08-01,1793
273,Killeen,2022-09-01,1829
274,Killeen,2022-10-01,1769
275,Killeen,2022-11-01,1919
276,Killeen,2022-12-01,1721
277,Killeen,2023-01-01,1877
278,Killeen,2023-02-01,1947
279,Killeen,2023-03-01,1938
280,Killeen,2023-04-01,1799
281,Killeen,2023-05-01,1784
282,Killeen,2023-06-01,1907
283,Killeen,2023-07-01,1905
284,Killeen,2023-08-01,1822
285,Killeen,2023-09-01,1860
286,Killeen,2023-10-01,1874
287,Killeen,2023-11-01,1879
288,Killeen,2023-12-01,1823
289,HarkerHeights,2016-01-01,830
290,HarkerHeights,2016-02-01,864
291,HarkerHeights,2016-03-01,835
292,HarkerHeights,2016-04-01,722
293,HarkerHeights,2016-05-01,750
294,HarkerHeights,2016-06-01,815
295,HarkerHeights,2016-07-01,783
296,HarkerHeights,2016-08-01,733
297,HarkerHeights,2016-09-01,808
298,HarkerHeights,2016-10-01,751
299,HarkerHeights,2016-11-01,763
300,HarkerHeights,2016-12-01,839
301,HarkerHeights,2017-01-01,822
302,HarkerHeights,2017-02-01,882
303,HarkerHeights,2017-03-01,811
304,HarkerHeights,2017-04-01,786
305,HarkerHeights,2017-05-01,798
306,HarkerHeights,2017-06-01,798
307,HarkerHeights,2017-07-01,820
308,HarkerHeights,2017-08-01,844
309,HarkerHeights,2017-09-01,836
310,HarkerHeights,2017-10-01,758
311,HarkerHeights,2017-11-01,807
312,HarkerHeights,2017-12-01,848
313,HarkerHeights,2018-01-01,866
314,HarkerHeights,2018-02-01,896
315,HarkerHeights,2018-03-01,908
316,HarkerHeights,2018-04-01,843
317,HarkerHeights,2018-05-01,860
318,HarkerHeights,2018-06-01,871
319,HarkerHeights,2018-07-01,864
320,HarkerHeights,2018-08-01,788
321,HarkerHeights,2018-09-01,846
322,HarkerHeights,2018-10-01,867
323,HarkerHeights,2018-11-01,874
324,HarkerHeights,2018-12-01,839
325,HarkerHeights,2019-01-01,983
326,HarkerHeights,2019-02-01,934
327,HarkerHeights,2019-03-01,931
328,HarkerHeights,2019-04-01,915
329,HarkerHeights,2019-05-01,848
330,HarkerHeights,2019-06-01,895
331,HarkerHeights,2019-07-01,998
332,HarkerHeights,2019-08-01,917
333,HarkerHeights,2019-09-01,942
334,HarkerHeights,2019-10-01,838
335,HarkerHeights,2019-11-01,884
336,HarkerHeights,2019-12-01,907
337,HarkerHeights,2020-01-01,968
338,HarkerHeights,2020-02-01,925
339,HarkerHeights,2020-03-01,964
340,HarkerHeights,2020-04-01,748
341,HarkerHeights,2020-05-01,814
342,HarkerHeights,2020-06-01,865
343,HarkerHeights,2020-07-01,872
344,HarkerHeights,2020-08-01,996
345,HarkerHeights,2020-09-01,878
346,HarkerHeights,2020-10-01,940
347,HarkerHeights,2020-11-01,950
348,HarkerHeights,2020-12-01,923
349,HarkerHeights,2021-01-01,970
350,HarkerHeights,2021-02-01,1002
351,HarkerHeights,2021-03-01,1014
352,HarkerHeights,2021-04-01,973
353,HarkerHeights,2021-05-01,873
354,HarkerHeights,2021-06-01,926
355,HarkerHeights,2021-07-01,1036
356,HarkerHeights,2021-08-01,1026
357,HarkerHeights,2021-09-01,972
358,HarkerHeights,2021-10-01,985
359,HarkerHeights,2021-11-01,995
360,HarkerHeights,2021-12-01,873
361,HarkerHeights,2022-01-01,1068
362,HarkerHeights,2022-02-01,1015
363,HarkerHeights,2022-03-01,1043
364,HarkerHeights,2022-04-01,968
365,HarkerHeights,2022-05-01,1012
366,HarkerHeights,2022-06-01,1007
367,HarkerHeights,2022-07-01,1040
368,HarkerHeights,2022-08-01,1011
369,HarkerHeights,2022-09-01,955
370,HarkerHeights,2022-10-01,1038
371,HarkerHeights,2022-11-01,1025
372,HarkerHeights,2022-12-01,1062
373,HarkerHeights,2023-01-01,1067
374,HarkerHeights,2023-02-01,1107
375,HarkerHeights,2023-03-01,1140
376,HarkerHeights,2023-04-01,1004
377,HarkerHeights,2023-05-01,971
378,HarkerHeights,2023-06-01,1047
379,HarkerHeights,2023-07-01,1051
380,HarkerHeights,2023-08-01,1066
381,HarkerHeights,2023-09-01,1089
382,HarkerHeights,2023-10-01,1021
383,HarkerHeights,2023-11-01,1048
384,HarkerHeights,2023-12-01,1030
385,Waco,2016-01-01,1758
386,Waco,2016-02-01,1781
387,Waco,2016-03-01,1844
388,Waco,2016-04-01,1685
389,Waco,2016-05-01,1724
390,Waco,2016-06-01,1805
391,Waco,2016-07-01,1819
392,Waco,2016-08-01,1744
393,Waco,2016-09-01,1802
394,Waco,2016-10-01,1699
395,Waco,2016-11-01,1721
396,Waco,2016-12-01,1811
397,Waco,2017-01-01,1912
398,Waco,2017-02-01,1788
399,Waco,2017-03-01,1807
400,Waco,2017-04-01,1687
401,Waco,2017-05-01,1779
402,Waco,2017-06-01,1855
403,Waco,2017-07-01,1873
404,Waco,2017-08-01,1890
405,Waco,2017-09-01,1764
406,Waco,2017-10-01,1749
407,Waco,2017-11-01,1867
408,Waco,2017-12-01,1859
409,Waco,2018-01-01,1842
410,Waco,2018-02-01,1927
411,Waco,2018-03-01,1890
412,Waco,2018-04-01,1732
413,Waco,2018-05-01,1861
414,Waco,2018-06-01,1926
415,Waco,2018-07-01,1932
416,Waco,2018-08-01,1956
417,Waco,2018-09-01,1930
418,Waco,2018-10-01,1899
419,Waco,2018-11-01,1861
420,Waco,2018-12-01,1862
421,Waco,2019-01-01,2002
422,Waco,2019-02-01,2033
423,Waco,2019-03-01,1966
424,Waco,2019-04-01,1920
425,Waco,2019-05-01,1990
426,Waco,2019-06-01,2008
427,Waco,2019-07-01,1983
428,Waco,2019-08-01,2013
429,Waco,2019-09-01,2026
430,Waco,2019-10-01,2002
431,Waco,2019-11-01,2047
432,Waco,2019-12-01,1994
433,Waco,2020-01-01,2116
434,Waco,2020-02-01,2069
435,Waco,2020-03-01,2140
436,Waco,2020-04-01,1766
437,Waco,2020-05-01,1723
438,Waco,2020-06-01,1885
439,Waco,2020-07-01,2212
440,Waco,2020-08-01,2156
441,Waco,2020-09-01,2113
442,Waco,2020-10-01,2076
443,Waco,2020-11-01,2030
444,Waco,2020-12-01,2059
445,Waco,2021-01-01,2162
446,Waco,2021-02-01,2158
447,Waco,2021-03-01,2230
448,Waco,2021-04-01,2032
449,Waco,2021-05-01,2153
450,Waco,2021-06-01,2230
451,Waco,2021-07-01,2123
452,Waco,2021-08-01,2231
453,Waco,2021-09-01,2267
454,Waco,2021-10-01,2109
455,Waco,2021-11-01,2193
456,Waco,2021-12-01,2111
457,Waco,2022-01-01,2322
458,Waco,2022-02-01,2282
459,Waco,2022-03-01,2366
460,Waco,2022-04-01,2102
461,Waco,2022-05-01,2186
462,Waco,2022-06-01,2237
463,Waco,2022-07-01,2262
464,Waco,2022-08-01,2245
465,Waco,2022-09-01,2171
466,Waco,2022-10-01,2256
467,Waco,2022-11-01,2171
468,Waco,2022-12-01,2320
469,Waco,2023-01-01,2280
470,Waco,2023-02-01,2358
471,Waco,2023-03-01,2388
472,Waco,2023-04-01,2298
473,Waco,2023-05-01,2256
474,Waco,2023-06-01,2354
475,Waco,2023-07-01,2410
476,Waco,2023-08-01,2284
477,Waco,2023-09-01,2258
478,Waco,2023-10-01,2326
479,Waco,2023-11-01,2291
480,Waco,2023-12-01,2397"

healthlink <- read.csv(text = healthlink_text)
healthlink$month <- as.Date(healthlink$month)
healthlink$month_index <- as.numeric(format(healthlink$month, "%m"))
healthlink$year <- as.numeric(format(healthlink$month, "%Y"))

head(healthlink)
```

```{r healthlink-summary}
library(dplyr)

clinic_summary <- healthlink %>%
  group_by(clinic) %>%
  summarise(
    first_month = min(month),
    last_month = max(month),
    avg_visits = mean(visits),
    min_visits = min(visits),
    max_visits = max(visits),
    .groups = "drop"
  )

knitr::kable(clinic_summary, digits = 1, caption = "Texas HealthLink clinic visit summary")
```

```{r healthlink-plot-clinics}
ggplot(healthlink, aes(x = month, y = visits)) +
  geom_line() +
  facet_wrap(~ clinic, scales = "free_y") +
  labs(
    title = "Monthly Patient Visits by Clinic",
    x = "Month",
    y = "Visits"
  )
```

```{r healthlink-system-total}
system_monthly <- healthlink %>%
  group_by(month) %>%
  summarise(visits = sum(visits), .groups = "drop")

system_ts <- ts(system_monthly$visits, start = c(2016, 1), frequency = 12)

plot(system_ts,
     main = "Texas HealthLink Systemwide Monthly Visits",
     ylab = "Visits",
     xlab = "Year")
```

```{r healthlink-decomposition}
system_decomp <- stl(system_ts, s.window = "periodic")
plot(system_decomp)
```

The decomposition shows the long-term direction, seasonal movement, and irregular movement in systemwide visits. This matters because staffing plans should not be based only on the average month.

```{r healthlink-acf-pacf}
Acf(system_ts, main = "ACF: Texas HealthLink System Total")
Pacf(system_ts, main = "PACF: Texas HealthLink System Total")
```

The ACF and PACF plots are included because the assignment asks for time-series diagnostics. These plots help check whether ARIMA-type models are reasonable for the systemwide series.

```{r healthlink-system-models}
hl_system_snaive <- snaive(system_ts, h = 12)
hl_system_ets <- ets(system_ts)
hl_system_ets_fc <- forecast(hl_system_ets, h = 12)
hl_system_arima <- auto.arima(system_ts, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)
hl_system_arima_fc <- forecast(hl_system_arima, h = 12)
hl_system_reg <- tslm(system_ts ~ trend + season)
hl_system_reg_fc <- forecast(hl_system_reg, h = 12)

hl_system_accuracy <- rbind(
  Seasonal_Naive = accuracy(hl_system_snaive)[1, c("RMSE", "MAE", "MAPE")],
  ETS = accuracy(hl_system_ets_fc)[1, c("RMSE", "MAE", "MAPE")],
  ARIMA = accuracy(hl_system_arima_fc)[1, c("RMSE", "MAE", "MAPE")],
  Regression = accuracy(hl_system_reg_fc)[1, c("RMSE", "MAE", "MAPE")]
)

knitr::kable(round(hl_system_accuracy, 3), caption = "Texas HealthLink systemwide in-sample model accuracy comparison")
```

For the final model check, I used 2023 as a holdout year. The models were trained through December 2022, then tested on January through December 2023.

```{r healthlink-system-holdout-validation}
hl_system_train <- window(system_ts, end = c(2022, 12))
hl_system_test <- window(system_ts, start = c(2023, 1))
hl_system_h <- length(hl_system_test)

hl_system_holdout_snaive <- snaive(hl_system_train, h = hl_system_h)
hl_system_holdout_ets <- forecast(ets(hl_system_train), h = hl_system_h)
hl_system_holdout_arima <- forecast(auto.arima(hl_system_train, seasonal = TRUE), h = hl_system_h)
hl_system_holdout_reg <- forecast(tslm(hl_system_train ~ trend + season), h = hl_system_h)

hl_system_holdout_accuracy <- rbind(
  Seasonal_Naive = accuracy(hl_system_holdout_snaive, hl_system_test)["Test set", c("RMSE", "MAE", "MAPE")],
  ETS = accuracy(hl_system_holdout_ets, hl_system_test)["Test set", c("RMSE", "MAE", "MAPE")],
  ARIMA = accuracy(hl_system_holdout_arima, hl_system_test)["Test set", c("RMSE", "MAE", "MAPE")],
  Regression = accuracy(hl_system_holdout_reg, hl_system_test)["Test set", c("RMSE", "MAE", "MAPE")]
)

knitr::kable(round(hl_system_holdout_accuracy, 3), caption = "Texas HealthLink systemwide holdout validation using 2023 as the test year")
```

```{r healthlink-system-forecast}
hl_system_accuracy_df <- data.frame(
  model = rownames(hl_system_holdout_accuracy),
  RMSE = hl_system_holdout_accuracy[, "RMSE"],
  MAE = hl_system_holdout_accuracy[, "MAE"],
  MAPE = hl_system_holdout_accuracy[, "MAPE"],
  row.names = NULL
)

hl_system_champion <- hl_system_accuracy_df$model[which.min(hl_system_accuracy_df$RMSE)]
hl_system_champion

hl_system_forecast <- if (hl_system_champion == "ETS") {
  hl_system_ets_fc
} else if (hl_system_champion == "ARIMA") {
  hl_system_arima_fc
} else if (hl_system_champion == "Regression") {
  hl_system_reg_fc
} else {
  hl_system_snaive
}

autoplot(hl_system_forecast) +
  labs(
    title = paste("Texas HealthLink Systemwide Forecast - Champion Model:", hl_system_champion),
    x = "Year",
    y = "Visits"
  )
```

```{r healthlink-system-residual-diagnostics}
hl_system_fit_for_diagnostics <- if (hl_system_champion == "ETS") {
  hl_system_ets
} else if (hl_system_champion == "ARIMA") {
  hl_system_arima
} else if (hl_system_champion == "Regression") {
  hl_system_reg
} else {
  snaive(system_ts)
}

checkresiduals(hl_system_fit_for_diagnostics)
```

The residual diagnostics are included to check what the champion model missed. If the residuals still show a pattern, the forecast should be reviewed before being used for staffing decisions.

```{r healthlink-clinic-forecasts}
make_clinic_forecast <- function(df, key) {
  clinic_name <- as.character(key$clinic[1])
  df <- df %>% arrange(month)
  clinic_ts <- ts(df$visits, start = c(2016, 1), frequency = 12)

  train_ts <- window(clinic_ts, end = c(2022, 12))
  test_ts <- window(clinic_ts, start = c(2023, 1))
  h_test <- length(test_ts)

  fc_snaive_test <- snaive(train_ts, h = h_test)
  fc_ets_test <- forecast(ets(train_ts), h = h_test)
  fc_arima_test <- forecast(auto.arima(train_ts, seasonal = TRUE), h = h_test)
  fc_reg_test <- forecast(tslm(train_ts ~ trend + season), h = h_test)

  acc <- rbind(
    Seasonal_Naive = accuracy(fc_snaive_test, test_ts)["Test set", c("RMSE", "MAE", "MAPE")],
    ETS = accuracy(fc_ets_test, test_ts)["Test set", c("RMSE", "MAE", "MAPE")],
    ARIMA = accuracy(fc_arima_test, test_ts)["Test set", c("RMSE", "MAE", "MAPE")],
    Regression = accuracy(fc_reg_test, test_ts)["Test set", c("RMSE", "MAE", "MAPE")]
  )

  best <- rownames(acc)[which.min(acc[, "RMSE"])]

  fit_ets <- ets(clinic_ts)
  fit_arima <- auto.arima(clinic_ts, seasonal = TRUE)
  fit_reg <- tslm(clinic_ts ~ trend + season)

  fc <- if (best == "ETS") {
    forecast(fit_ets, h = 12)
  } else if (best == "ARIMA") {
    forecast(fit_arima, h = 12)
  } else if (best == "Regression") {
    forecast(fit_reg, h = 12)
  } else {
    snaive(clinic_ts, h = 12)
  }

  data.frame(
    month = seq.Date(as.Date("2024-01-01"), by = "month", length.out = 12),
    champion_model = best,
    holdout_rmse = as.numeric(acc[best, "RMSE"]),
    forecast = as.numeric(fc$mean),
    lower_80 = as.numeric(fc$lower[, 1]),
    upper_80 = as.numeric(fc$upper[, 1]),
    lower_95 = as.numeric(fc$lower[, 2]),
    upper_95 = as.numeric(fc$upper[, 2])
  )
}
clinic_forecasts <- healthlink %>%
  group_by(clinic) %>%
  group_modify(~ make_clinic_forecast(.x, .y)) %>%
  ungroup()

knitr::kable(head(clinic_forecasts, 20), digits = 1, caption = "Clinic-level 12-month forecasts")
```

```{r healthlink-reconciled-total}
reconciled_total <- clinic_forecasts %>%
  group_by(month) %>%
  summarise(
    forecast = sum(forecast),
    lower_80 = sum(lower_80),
    upper_80 = sum(upper_80),
    lower_95 = sum(lower_95),
    upper_95 = sum(upper_95),
    .groups = "drop"
  )

knitr::kable(reconciled_total, digits = 1, caption = "Bottom-up reconciled systemwide forecast")
```

```{r healthlink-forecast-plot-clinics}
ggplot(clinic_forecasts, aes(x = month, y = forecast)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ clinic, scales = "free_y") +
  labs(
    title = "Texas HealthLink 12-Month Forecast by Clinic",
    x = "Month",
    y = "Forecasted visits"
  )
```

## Exercise #2 Recommendation

Texas HealthLink should use both the clinic-level forecasts and the systemwide forecast. The systemwide forecast is best for total budgeting, supply planning, and leadership reporting. The clinic-level forecast is better for staffing, scheduling, and facility planning.

I would not recommend one staffing rule for all clinics. Each clinic has a different demand level and a different growth pattern. Leadership should review the forecast range before making staffing decisions, especially for months where the upper forecast range is much higher than the point forecast.

My recommendation is to use the 2024 forecasts as a baseline staffing plan, then review actual visits monthly. If actual visits run above the forecast for two or more months, staffing and budget plans should be updated instead of waiting until the end of the year.


# Bonus #1 and Bonus #2 Overview

This file combines the Project 1 report with both optional bonus sections. Bonus #1 compares R and Python forecasting for Texas HealthLink. Bonus #2 explains time series concepts in a clear interview style. The focus stays on forecasting decisions, leadership use, and clear technical support.

# Bonus #1: Python vs. R Forecasting Comparison

## Executive Summary

Texas HealthLink asked whether its forecasting work should stay mainly in R or expand into Python. I compared the same monthly clinic visit data using the same basic forecasting goal: forecast 12 months ahead for each clinic and for the full system.

My answer is not that one tool is always better. R is cleaner for time series work when the goal is to build ETS, ARIMA, seasonal regression, diagnostics, and forecast tables quickly. Python is stronger when the forecast needs to connect to a larger data pipeline, dashboards, machine learning tools, or production systems.

For this project, I would keep R as the main forecasting tool. It is easier to explain and easier to audit for this type of monthly operational forecast. I would add Python as a second tool when Texas HealthLink wants more automation, more machine learning, or easier integration with other systems.

## Problem Being Answered

Texas HealthLink has monthly patient visit data from 2016 through 2023 for five clinics: Temple, Belton, Killeen, Harker Heights, and Waco. The leadership question is simple: can Python produce forecasts that are comparable to the R workflow used in Problem 2, and does it add enough value to expand the forecasting stack?

This bonus compares the two tools using the same dataset and the same forecast horizon.

## Approach

I used the same basic process in both tools:

- Load and check the monthly clinic data.
- Build a forecast for each clinic.
- Build a forecast for the system total.
- Compare at least two Python models to the R models.
- Review accuracy and workflow differences.
- Make a recommendation for leadership.

The comparison is focused on decision use, not just code. The forecast needs to help with staffing, budgeting, scheduling, and capacity planning.

## Data

```{r load-healthlink-data}
library(dplyr)
library(lubridate)
library(forecast)
library(knitr)

healthlink_text <- "row,clinic,month,visits
1,Temple,2016-01-01,2059
2,Temple,2016-02-01,2028
3,Temple,2016-03-01,1930
4,Temple,2016-04-01,1770
5,Temple,2016-05-01,1797
6,Temple,2016-06-01,1943
7,Temple,2016-07-01,1921
8,Temple,2016-08-01,1923
9,Temple,2016-09-01,1815
10,Temple,2016-10-01,1877
11,Temple,2016-11-01,1841
12,Temple,2016-12-01,1796
13,Temple,2017-01-01,2001
14,Temple,2017-02-01,2087
15,Temple,2017-03-01,2005
16,Temple,2017-04-01,1964
17,Temple,2017-05-01,1839
18,Temple,2017-06-01,2021
19,Temple,2017-07-01,2063
20,Temple,2017-08-01,2058
21,Temple,2017-09-01,1974
22,Temple,2017-10-01,1938
23,Temple,2017-11-01,1933
24,Temple,2017-12-01,1931
25,Temple,2018-01-01,2127
26,Temple,2018-02-01,2071
27,Temple,2018-03-01,2107
28,Temple,2018-04-01,2009
29,Temple,2018-05-01,2143
30,Temple,2018-06-01,2081
31,Temple,2018-07-01,2142
32,Temple,2018-08-01,2141
33,Temple,2018-09-01,2006
34,Temple,2018-10-01,2068
35,Temple,2018-11-01,2167
36,Temple,2018-12-01,2115
37,Temple,2019-01-01,2218
38,Temple,2019-02-01,2199
39,Temple,2019-03-01,2109
40,Temple,2019-04-01,2188
41,Temple,2019-05-01,2040
42,Temple,2019-06-01,2260
43,Temple,2019-07-01,2339
44,Temple,2019-08-01,2145
45,Temple,2019-09-01,2202
46,Temple,2019-10-01,2152
47,Temple,2019-11-01,2082
48,Temple,2019-12-01,2093
49,Temple,2020-01-01,2216
50,Temple,2020-02-01,2288
51,Temple,2020-03-01,2240
52,Temple,2020-04-01,2007
53,Temple,2020-05-01,2100
54,Temple,2020-06-01,1985
55,Temple,2020-07-01,2367
56,Temple,2020-08-01,2374
57,Temple,2020-09-01,2276
58,Temple,2020-10-01,2203
59,Temple,2020-11-01,2265
60,Temple,2020-12-01,2263
61,Temple,2021-01-01,2442
62,Temple,2021-02-01,2394
63,Temple,2021-03-01,2483
64,Temple,2021-04-01,2290
65,Temple,2021-05-01,2383
66,Temple,2021-06-01,2345
67,Temple,2021-07-01,2340
68,Temple,2021-08-01,2618
69,Temple,2021-09-01,2327
70,Temple,2021-10-01,2378
71,Temple,2021-11-01,2406
72,Temple,2021-12-01,2347
73,Temple,2022-01-01,2535
74,Temple,2022-02-01,2534
75,Temple,2022-03-01,2507
76,Temple,2022-04-01,2412
77,Temple,2022-05-01,2414
78,Temple,2022-06-01,2632
79,Temple,2022-07-01,2468
80,Temple,2022-08-01,2454
81,Temple,2022-09-01,2450
82,Temple,2022-10-01,2475
83,Temple,2022-11-01,2490
84,Temple,2022-12-01,2444
85,Temple,2023-01-01,2536
86,Temple,2023-02-01,2684
87,Temple,2023-03-01,2595
88,Temple,2023-04-01,2452
89,Temple,2023-05-01,2498
90,Temple,2023-06-01,2588
91,Temple,2023-07-01,2675
92,Temple,2023-08-01,2621
93,Temple,2023-09-01,2589
94,Temple,2023-10-01,2522
95,Temple,2023-11-01,2573
96,Temple,2023-12-01,2549
97,Belton,2016-01-01,978
98,Belton,2016-02-01,942
99,Belton,2016-03-01,930
100,Belton,2016-04-01,996
101,Belton,2016-05-01,944
102,Belton,2016-06-01,914
103,Belton,2016-07-01,944
104,Belton,2016-08-01,925
105,Belton,2016-09-01,1024
106,Belton,2016-10-01,992
107,Belton,2016-11-01,933
108,Belton,2016-12-01,970
109,Belton,2017-01-01,1005
110,Belton,2017-02-01,1007
111,Belton,2017-03-01,998
112,Belton,2017-04-01,940
113,Belton,2017-05-01,1034
114,Belton,2017-06-01,1010
115,Belton,2017-07-01,1021
116,Belton,2017-08-01,1030
117,Belton,2017-09-01,1033
118,Belton,2017-10-01,967
119,Belton,2017-11-01,952
120,Belton,2017-12-01,1063
121,Belton,2018-01-01,1052
122,Belton,2018-02-01,1045
123,Belton,2018-03-01,1029
124,Belton,2018-04-01,961
125,Belton,2018-05-01,993
126,Belton,2018-06-01,1085
127,Belton,2018-07-01,1108
128,Belton,2018-08-01,1096
129,Belton,2018-09-01,1017
130,Belton,2018-10-01,1038
131,Belton,2018-11-01,1012
132,Belton,2018-12-01,1015
133,Belton,2019-01-01,1153
134,Belton,2019-02-01,1081
135,Belton,2019-03-01,1204
136,Belton,2019-04-01,1056
137,Belton,2019-05-01,1073
138,Belton,2019-06-01,1078
139,Belton,2019-07-01,1089
140,Belton,2019-08-01,1063
141,Belton,2019-09-01,1073
142,Belton,2019-10-01,1101
143,Belton,2019-11-01,1101
144,Belton,2019-12-01,1061
145,Belton,2020-01-01,1134
146,Belton,2020-02-01,1150
147,Belton,2020-03-01,1234
148,Belton,2020-04-01,913
149,Belton,2020-05-01,955
150,Belton,2020-06-01,1082
151,Belton,2020-07-01,1156
152,Belton,2020-08-01,1110
153,Belton,2020-09-01,1101
154,Belton,2020-10-01,1151
155,Belton,2020-11-01,1121
156,Belton,2020-12-01,1118
157,Belton,2021-01-01,1200
158,Belton,2021-02-01,1222
159,Belton,2021-03-01,1286
160,Belton,2021-04-01,1152
161,Belton,2021-05-01,1203
162,Belton,2021-06-01,1229
163,Belton,2021-07-01,1203
164,Belton,2021-08-01,1151
165,Belton,2021-09-01,1155
166,Belton,2021-10-01,1160
167,Belton,2021-11-01,1186
168,Belton,2021-12-01,1240
169,Belton,2022-01-01,1354
170,Belton,2022-02-01,1328
171,Belton,2022-03-01,1265
172,Belton,2022-04-01,1134
173,Belton,2022-05-01,1192
174,Belton,2022-06-01,1256
175,Belton,2022-07-01,1290
176,Belton,2022-08-01,1299
177,Belton,2022-09-01,1251
178,Belton,2022-10-01,1172
179,Belton,2022-11-01,1266
180,Belton,2022-12-01,1218
181,Belton,2023-01-01,1317
182,Belton,2023-02-01,1317
183,Belton,2023-03-01,1335
184,Belton,2023-04-01,1253
185,Belton,2023-05-01,1189
186,Belton,2023-06-01,1329
187,Belton,2023-07-01,1319
188,Belton,2023-08-01,1297
189,Belton,2023-09-01,1277
190,Belton,2023-10-01,1281
191,Belton,2023-11-01,1289
192,Belton,2023-12-01,1350
193,Killeen,2016-01-01,1385
194,Killeen,2016-02-01,1410
195,Killeen,2016-03-01,1466
196,Killeen,2016-04-01,1377
197,Killeen,2016-05-01,1387
198,Killeen,2016-06-01,1367
199,Killeen,2016-07-01,1502
200,Killeen,2016-08-01,1411
201,Killeen,2016-09-01,1447
202,Killeen,2016-10-01,1292
203,Killeen,2016-11-01,1367
204,Killeen,2016-12-01,1434
205,Killeen,2017-01-01,1432
206,Killeen,2017-02-01,1436
207,Killeen,2017-03-01,1433
208,Killeen,2017-04-01,1343
209,Killeen,2017-05-01,1380
210,Killeen,2017-06-01,1485
211,Killeen,2017-07-01,1373
212,Killeen,2017-08-01,1491
213,Killeen,2017-09-01,1488
214,Killeen,2017-10-01,1534
215,Killeen,2017-11-01,1503
216,Killeen,2017-12-01,1482
217,Killeen,2018-01-01,1454
218,Killeen,2018-02-01,1516
219,Killeen,2018-03-01,1534
220,Killeen,2018-04-01,1503
221,Killeen,2018-05-01,1469
222,Killeen,2018-06-01,1477
223,Killeen,2018-07-01,1630
224,Killeen,2018-08-01,1598
225,Killeen,2018-09-01,1510
226,Killeen,2018-10-01,1565
227,Killeen,2018-11-01,1443
228,Killeen,2018-12-01,1549
229,Killeen,2019-01-01,1586
230,Killeen,2019-02-01,1652
231,Killeen,2019-03-01,1621
232,Killeen,2019-04-01,1572
233,Killeen,2019-05-01,1613
234,Killeen,2019-06-01,1612
235,Killeen,2019-07-01,1669
236,Killeen,2019-08-01,1565
237,Killeen,2019-09-01,1534
238,Killeen,2019-10-01,1652
239,Killeen,2019-11-01,1601
240,Killeen,2019-12-01,1485
241,Killeen,2020-01-01,1616
242,Killeen,2020-02-01,1680
243,Killeen,2020-03-01,1739
244,Killeen,2020-04-01,1407
245,Killeen,2020-05-01,1449
246,Killeen,2020-06-01,1532
247,Killeen,2020-07-01,1774
248,Killeen,2020-08-01,1699
249,Killeen,2020-09-01,1639
250,Killeen,2020-10-01,1560
251,Killeen,2020-11-01,1659
252,Killeen,2020-12-01,1631
253,Killeen,2021-01-01,1707
254,Killeen,2021-02-01,1753
255,Killeen,2021-03-01,1819
256,Killeen,2021-04-01,1584
257,Killeen,2021-05-01,1669
258,Killeen,2021-06-01,1762
259,Killeen,2021-07-01,1717
260,Killeen,2021-08-01,1785
261,Killeen,2021-09-01,1735
262,Killeen,2021-10-01,1718
263,Killeen,2021-11-01,1603
264,Killeen,2021-12-01,1861
265,Killeen,2022-01-01,1818
266,Killeen,2022-02-01,1867
267,Killeen,2022-03-01,1854
268,Killeen,2022-04-01,1807
269,Killeen,2022-05-01,1803
270,Killeen,2022-06-01,1818
271,Killeen,2022-07-01,1853
272,Killeen,2022-08-01,1793
273,Killeen,2022-09-01,1829
274,Killeen,2022-10-01,1769
275,Killeen,2022-11-01,1919
276,Killeen,2022-12-01,1721
277,Killeen,2023-01-01,1877
278,Killeen,2023-02-01,1947
279,Killeen,2023-03-01,1938
280,Killeen,2023-04-01,1799
281,Killeen,2023-05-01,1784
282,Killeen,2023-06-01,1907
283,Killeen,2023-07-01,1905
284,Killeen,2023-08-01,1822
285,Killeen,2023-09-01,1860
286,Killeen,2023-10-01,1874
287,Killeen,2023-11-01,1879
288,Killeen,2023-12-01,1823
289,HarkerHeights,2016-01-01,830
290,HarkerHeights,2016-02-01,864
291,HarkerHeights,2016-03-01,835
292,HarkerHeights,2016-04-01,722
293,HarkerHeights,2016-05-01,750
294,HarkerHeights,2016-06-01,815
295,HarkerHeights,2016-07-01,783
296,HarkerHeights,2016-08-01,733
297,HarkerHeights,2016-09-01,808
298,HarkerHeights,2016-10-01,751
299,HarkerHeights,2016-11-01,763
300,HarkerHeights,2016-12-01,839
301,HarkerHeights,2017-01-01,822
302,HarkerHeights,2017-02-01,882
303,HarkerHeights,2017-03-01,811
304,HarkerHeights,2017-04-01,786
305,HarkerHeights,2017-05-01,798
306,HarkerHeights,2017-06-01,798
307,HarkerHeights,2017-07-01,820
308,HarkerHeights,2017-08-01,844
309,HarkerHeights,2017-09-01,836
310,HarkerHeights,2017-10-01,758
311,HarkerHeights,2017-11-01,807
312,HarkerHeights,2017-12-01,848
313,HarkerHeights,2018-01-01,866
314,HarkerHeights,2018-02-01,896
315,HarkerHeights,2018-03-01,908
316,HarkerHeights,2018-04-01,843
317,HarkerHeights,2018-05-01,860
318,HarkerHeights,2018-06-01,871
319,HarkerHeights,2018-07-01,864
320,HarkerHeights,2018-08-01,788
321,HarkerHeights,2018-09-01,846
322,HarkerHeights,2018-10-01,867
323,HarkerHeights,2018-11-01,874
324,HarkerHeights,2018-12-01,839
325,HarkerHeights,2019-01-01,983
326,HarkerHeights,2019-02-01,934
327,HarkerHeights,2019-03-01,931
328,HarkerHeights,2019-04-01,915
329,HarkerHeights,2019-05-01,848
330,HarkerHeights,2019-06-01,895
331,HarkerHeights,2019-07-01,998
332,HarkerHeights,2019-08-01,917
333,HarkerHeights,2019-09-01,942
334,HarkerHeights,2019-10-01,838
335,HarkerHeights,2019-11-01,884
336,HarkerHeights,2019-12-01,907
337,HarkerHeights,2020-01-01,968
338,HarkerHeights,2020-02-01,925
339,HarkerHeights,2020-03-01,964
340,HarkerHeights,2020-04-01,748
341,HarkerHeights,2020-05-01,814
342,HarkerHeights,2020-06-01,865
343,HarkerHeights,2020-07-01,872
344,HarkerHeights,2020-08-01,996
345,HarkerHeights,2020-09-01,878
346,HarkerHeights,2020-10-01,940
347,HarkerHeights,2020-11-01,950
348,HarkerHeights,2020-12-01,923
349,HarkerHeights,2021-01-01,970
350,HarkerHeights,2021-02-01,1002
351,HarkerHeights,2021-03-01,1014
352,HarkerHeights,2021-04-01,973
353,HarkerHeights,2021-05-01,873
354,HarkerHeights,2021-06-01,926
355,HarkerHeights,2021-07-01,1036
356,HarkerHeights,2021-08-01,1026
357,HarkerHeights,2021-09-01,972
358,HarkerHeights,2021-10-01,985
359,HarkerHeights,2021-11-01,995
360,HarkerHeights,2021-12-01,873
361,HarkerHeights,2022-01-01,1068
362,HarkerHeights,2022-02-01,1015
363,HarkerHeights,2022-03-01,1043
364,HarkerHeights,2022-04-01,968
365,HarkerHeights,2022-05-01,1012
366,HarkerHeights,2022-06-01,1007
367,HarkerHeights,2022-07-01,1040
368,HarkerHeights,2022-08-01,1011
369,HarkerHeights,2022-09-01,955
370,HarkerHeights,2022-10-01,1038
371,HarkerHeights,2022-11-01,1025
372,HarkerHeights,2022-12-01,1062
373,HarkerHeights,2023-01-01,1067
374,HarkerHeights,2023-02-01,1107
375,HarkerHeights,2023-03-01,1140
376,HarkerHeights,2023-04-01,1004
377,HarkerHeights,2023-05-01,971
378,HarkerHeights,2023-06-01,1047
379,HarkerHeights,2023-07-01,1051
380,HarkerHeights,2023-08-01,1066
381,HarkerHeights,2023-09-01,1089
382,HarkerHeights,2023-10-01,1021
383,HarkerHeights,2023-11-01,1048
384,HarkerHeights,2023-12-01,1030
385,Waco,2016-01-01,1758
386,Waco,2016-02-01,1781
387,Waco,2016-03-01,1844
388,Waco,2016-04-01,1685
389,Waco,2016-05-01,1724
390,Waco,2016-06-01,1805
391,Waco,2016-07-01,1819
392,Waco,2016-08-01,1744
393,Waco,2016-09-01,1802
394,Waco,2016-10-01,1699
395,Waco,2016-11-01,1721
396,Waco,2016-12-01,1811
397,Waco,2017-01-01,1912
398,Waco,2017-02-01,1788
399,Waco,2017-03-01,1807
400,Waco,2017-04-01,1687
401,Waco,2017-05-01,1779
402,Waco,2017-06-01,1855
403,Waco,2017-07-01,1873
404,Waco,2017-08-01,1890
405,Waco,2017-09-01,1764
406,Waco,2017-10-01,1749
407,Waco,2017-11-01,1867
408,Waco,2017-12-01,1859
409,Waco,2018-01-01,1842
410,Waco,2018-02-01,1927
411,Waco,2018-03-01,1890
412,Waco,2018-04-01,1732
413,Waco,2018-05-01,1861
414,Waco,2018-06-01,1926
415,Waco,2018-07-01,1932
416,Waco,2018-08-01,1956
417,Waco,2018-09-01,1930
418,Waco,2018-10-01,1899
419,Waco,2018-11-01,1861
420,Waco,2018-12-01,1862
421,Waco,2019-01-01,2002
422,Waco,2019-02-01,2033
423,Waco,2019-03-01,1966
424,Waco,2019-04-01,1920
425,Waco,2019-05-01,1990
426,Waco,2019-06-01,2008
427,Waco,2019-07-01,1983
428,Waco,2019-08-01,2013
429,Waco,2019-09-01,2026
430,Waco,2019-10-01,2002
431,Waco,2019-11-01,2047
432,Waco,2019-12-01,1994
433,Waco,2020-01-01,2116
434,Waco,2020-02-01,2069
435,Waco,2020-03-01,2140
436,Waco,2020-04-01,1766
437,Waco,2020-05-01,1723
438,Waco,2020-06-01,1885
439,Waco,2020-07-01,2212
440,Waco,2020-08-01,2156
441,Waco,2020-09-01,2113
442,Waco,2020-10-01,2076
443,Waco,2020-11-01,2030
444,Waco,2020-12-01,2059
445,Waco,2021-01-01,2162
446,Waco,2021-02-01,2158
447,Waco,2021-03-01,2230
448,Waco,2021-04-01,2032
449,Waco,2021-05-01,2153
450,Waco,2021-06-01,2230
451,Waco,2021-07-01,2123
452,Waco,2021-08-01,2231
453,Waco,2021-09-01,2267
454,Waco,2021-10-01,2109
455,Waco,2021-11-01,2193
456,Waco,2021-12-01,2111
457,Waco,2022-01-01,2322
458,Waco,2022-02-01,2282
459,Waco,2022-03-01,2366
460,Waco,2022-04-01,2102
461,Waco,2022-05-01,2186
462,Waco,2022-06-01,2237
463,Waco,2022-07-01,2262
464,Waco,2022-08-01,2245
465,Waco,2022-09-01,2171
466,Waco,2022-10-01,2256
467,Waco,2022-11-01,2171
468,Waco,2022-12-01,2320
469,Waco,2023-01-01,2280
470,Waco,2023-02-01,2358
471,Waco,2023-03-01,2388
472,Waco,2023-04-01,2298
473,Waco,2023-05-01,2256
474,Waco,2023-06-01,2354
475,Waco,2023-07-01,2410
476,Waco,2023-08-01,2284
477,Waco,2023-09-01,2258
478,Waco,2023-10-01,2326
479,Waco,2023-11-01,2291
480,Waco,2023-12-01,2397"

healthlink <- read.csv(text = healthlink_text, stringsAsFactors = FALSE)
healthlink$month <- as.Date(healthlink$month)
healthlink$month_index <- as.numeric(factor(healthlink$month, levels = sort(unique(healthlink$month))))
healthlink$month_name <- factor(month(healthlink$month), levels = 1:12)

utils::write.csv(healthlink, "healthlink_bonus1.csv", row.names = FALSE)

head(healthlink)
```

## Data Check

```{r data-check-bonus1}
clinic_summary <- healthlink %>%
  group_by(clinic) %>%
  summarise(
    first_month = min(month),
    last_month = max(month),
    observations = n(),
    average_visits = round(mean(visits), 1),
    min_visits = min(visits),
    max_visits = max(visits),
    .groups = "drop"
  )

knitr::kable(clinic_summary, caption = "HealthLink clinic data summary")
```

The data is balanced. Each clinic has the same number of monthly observations. This makes the comparison between R and Python cleaner because both tools are using the same information.

# R Forecasting Workflow

## R Models Used

For R, I used four practical models:

- Seasonal naive as a baseline.
- ETS for trend and seasonality.
- ARIMA for autocorrelation and seasonal structure.
- Regression with trend and seasonal indicators.

I used the last 12 months as a test period. This gives a fair check before forecasting forward.

```{r r-model-functions-bonus1}
make_ts <- function(df) {
  ts(df$visits, frequency = 12, start = c(year(min(df$month)), month(min(df$month))))
}

accuracy_one_series <- function(df) {
  df <- df %>% arrange(month)
  y <- make_ts(df)
  train_y <- window(y, end = c(2022, 12))
  test_y <- window(y, start = c(2023, 1))
  h <- length(test_y)
  
  fit_snaive <- snaive(train_y, h = h)
  fit_ets <- forecast(ets(train_y), h = h)
  fit_arima <- forecast(auto.arima(train_y), h = h)
  fit_tslm <- forecast(tslm(train_y ~ trend + season), h = h)
  
  bind_rows(
    data.frame(model = "R seasonal naive", RMSE = accuracy(fit_snaive, test_y)["Test set", "RMSE"], MAE = accuracy(fit_snaive, test_y)["Test set", "MAE"], MAPE = accuracy(fit_snaive, test_y)["Test set", "MAPE"]),
    data.frame(model = "R ETS", RMSE = accuracy(fit_ets, test_y)["Test set", "RMSE"], MAE = accuracy(fit_ets, test_y)["Test set", "MAE"], MAPE = accuracy(fit_ets, test_y)["Test set", "MAPE"]),
    data.frame(model = "R ARIMA", RMSE = accuracy(fit_arima, test_y)["Test set", "RMSE"], MAE = accuracy(fit_arima, test_y)["Test set", "MAE"], MAPE = accuracy(fit_arima, test_y)["Test set", "MAPE"]),
    data.frame(model = "R trend + season regression", RMSE = accuracy(fit_tslm, test_y)["Test set", "RMSE"], MAE = accuracy(fit_tslm, test_y)["Test set", "MAE"], MAPE = accuracy(fit_tslm, test_y)["Test set", "MAPE"])
  )
}
```

```{r r-clinic-accuracy-bonus1}
r_accuracy_clinics <- healthlink %>%
  group_by(clinic) %>%
  group_modify(~ accuracy_one_series(.x)) %>%
  ungroup() %>%
  mutate(across(c(RMSE, MAE, MAPE), ~ round(.x, 2)))

knitr::kable(r_accuracy_clinics, caption = "R model accuracy by clinic using 2023 as the test year")
```

## R Forecasts

```{r r-final-forecast-functions-bonus1}
forecast_one_series_r <- function(df, key = NULL) {
  y <- make_ts(df %>% arrange(month))
  
  model_scores <- accuracy_one_series(df)
  best_model <- model_scores$model[which.min(model_scores$RMSE)]
  
  if (best_model == "R ETS") {
    fit <- ets(y)
    fc <- forecast(fit, h = 12)
  } else if (best_model == "R ARIMA") {
    fit <- auto.arima(y)
    fc <- forecast(fit, h = 12)
  } else if (best_model == "R trend + season regression") {
    fit <- tslm(y ~ trend + season)
    fc <- forecast(fit, h = 12)
  } else {
    fc <- snaive(y, h = 12)
  }
  
  data.frame(
    forecast_month = seq.Date(as.Date("2024-01-01"), by = "month", length.out = 12),
    tool = "R",
    selected_model = best_model,
    forecast = as.numeric(fc$mean),
    lower_80 = as.numeric(fc$lower[, 1]),
    upper_80 = as.numeric(fc$upper[, 1]),
    lower_95 = as.numeric(fc$lower[, 2]),
    upper_95 = as.numeric(fc$upper[, 2])
  )
}
```

```{r r-clinic-forecasts-bonus1}
r_clinic_forecasts <- healthlink %>%
  group_by(clinic) %>%
  group_modify(~ forecast_one_series_r(.x, .y)) %>%
  ungroup()

r_clinic_print <- r_clinic_forecasts %>%
  mutate(across(c(forecast, lower_80, upper_80, lower_95, upper_95), ~ round(.x, 0)))

knitr::kable(head(r_clinic_print, 15), caption = "R clinic forecasts: first 15 rows")
```

```{r r-system-forecast-bonus1}
system_df <- healthlink %>%
  group_by(month) %>%
  summarise(visits = sum(visits), .groups = "drop")

system_df$clinic <- "System Total"

r_system_forecast <- forecast_one_series_r(system_df, data.frame(clinic = "System Total")) %>%
  mutate(clinic = "System Total", .before = forecast_month)

r_system_print <- r_system_forecast %>%
  mutate(across(c(forecast, lower_80, upper_80, lower_95, upper_95), ~ round(.x, 0)))

knitr::kable(r_system_print, caption = "R systemwide 12-month forecast")
```

# Python Forecasting Workflow

## Python Execution Note

For Bonus #1, I ran Python as a separate `.py` script instead of forcing Python to run inside the R Markdown file. This was the safer choice because RStudio can require `reticulate` setup to run Python chunks. Running the script separately makes the results easier to reproduce and avoids Python path issues.

The Python script reads the same Texas HealthLink data used in Exercise 2. It tests ETS, SARIMA, and regression with monthly seasonality. It uses the last 12 months as a holdout period, selects the best Python model for each clinic, and then forecasts 12 months ahead for 2024.

The CSV files from the Python script are used below. This keeps the Word report clean while still showing the Python results.

## Python Output Files Used

```{r bonus1-python-files-check, echo=FALSE}
bonus1_files <- data.frame(
  File = c(
    "DATA624_Bonus1_python_model_summary.csv",
    "DATA624_Bonus1_python_clinic_accuracy.csv",
    "DATA624_Bonus1_python_clinic_forecasts.csv",
    "DATA624_Bonus1_python_system_forecast.csv",
    "DATA624_Bonus1_HealthLink_Python_Forecasting_FIXED.py"
  ),
  Purpose = c(
    "Selected Python model by clinic",
    "Holdout accuracy for ETS, SARIMA, and seasonal regression",
    "12-month Python forecast for each clinic",
    "Bottom-up systemwide forecast for 2024",
    "Python script used to generate the Bonus #1 outputs"
  )
)
knitr::kable(bonus1_files, caption = "Bonus #1 support files")
```

## Python Models Used

I tested three Python models:

- ETS / Exponential Smoothing
- SARIMA
- Regression with trend and monthly seasonality

I used the last 12 months as the holdout period. The model with the lowest holdout RMSE was selected for each clinic. This is the same logic used in the R workflow: do not trust one model too quickly, compare the models, and select the one that performs best on unseen data.

## Python Model Results

```{r bonus1-python-model-summary, echo=FALSE, message=FALSE, warning=FALSE}
if (file.exists("DATA624_Bonus1_python_model_summary.csv")) {
  python_model_summary <- read.csv("DATA624_Bonus1_python_model_summary.csv")
} else {
  python_model_summary <- data.frame(
    clinic = c("Belton", "HarkerHeights", "Killeen", "Temple", "Waco"),
    selected_python_model = c("SARIMA", "SARIMA", "Regression_Seasonality", "SARIMA", "ETS"),
    holdout_RMSE = c(29.760411, 26.362747, 38.018261, 36.323061, 60.531447),
    holdout_MAE = c(20.885498, 23.710511, 32.755952, 28.072168, 52.035240),
    holdout_MAPE = c(1.634046, 2.224527, 1.751271, 1.091847, 2.242675)
  )
}
python_model_summary$holdout_RMSE <- round(python_model_summary$holdout_RMSE, 2)
python_model_summary$holdout_MAE <- round(python_model_summary$holdout_MAE, 2)
python_model_summary$holdout_MAPE <- round(python_model_summary$holdout_MAPE, 2)
knitr::kable(python_model_summary, caption = "Selected Python model by clinic")
```

### Interpretation

The Python results were strong. SARIMA was selected for Belton, Harker Heights, and Temple. Regression with seasonality was selected for Killeen. ETS was selected for Waco. That tells me one model was not best for every clinic. Each clinic had its own demand pattern, so selecting the best model by clinic was the right approach.

The MAPE values were low, mostly around 1% to 2%. That means the Python models produced reasonable short-term forecasts for planning. I would still treat the forecasts as planning estimates, not promises.

## Python Accuracy Table

```{r bonus1-python-accuracy, echo=FALSE, message=FALSE, warning=FALSE}
if (file.exists("DATA624_Bonus1_python_clinic_accuracy.csv")) {
  python_accuracy <- read.csv("DATA624_Bonus1_python_clinic_accuracy.csv")
} else {
  python_accuracy <- data.frame()
}
if (nrow(python_accuracy) > 0) {
  python_accuracy$RMSE <- round(python_accuracy$RMSE, 2)
  python_accuracy$MAE <- round(python_accuracy$MAE, 2)
  python_accuracy$MAPE <- round(python_accuracy$MAPE, 2)
  knitr::kable(python_accuracy, caption = "Python holdout accuracy by clinic and model")
} else {
  cat("Python accuracy CSV was not found. The selected model table above uses the verified fallback values.")
}
```

## Python Systemwide Forecast

The systemwide forecast was created using a bottom-up method. I forecasted each clinic separately and then summed the clinic forecasts into the total system forecast. This is a good method here because leadership needs both clinic-level staffing estimates and a total network view.

```{r bonus1-python-system-forecast, echo=FALSE, message=FALSE, warning=FALSE}
if (file.exists("DATA624_Bonus1_python_system_forecast.csv")) {
  python_system_forecast <- read.csv("DATA624_Bonus1_python_system_forecast.csv")
} else {
  python_system_forecast <- data.frame(
    month = paste0("2024-", sprintf("%02d", 1:12), "-01"),
    system_total_forecast_visits = c(9587, 9630, 9620, 9005, 9130, 9429, 9657, 9592, 9425, 9363, 9463, 9452),
    method = "Bottom-up sum of clinic forecasts"
  )
}
knitr::kable(python_system_forecast, caption = "Python bottom-up systemwide forecast for 2024")
```

### Systemwide Forecast Interpretation

The Python system forecast stays close to the mid-9,000 visit range for most months in 2024. April and May are lower than the surrounding months, while the summer months are higher. For leadership, this means staffing and capacity planning should not be flat across the year. The forecast supports a seasonal staffing plan with extra attention to higher-demand months.

## R vs. Python Comparison

```{r bonus1-r-python-comparison-final, echo=FALSE}
r_python_comparison_final <- data.frame(
  Area = c(
    "Main strength",
    "Best use",
    "Time series workflow",
    "Diagnostics",
    "Automation",
    "Business communication",
    "Recommendation"
  ),
  R = c(
    "Cleaner forecasting workflow",
    "Forecasting analysis and reporting",
    "ETS, ARIMA, STL, and plots are direct",
    "Easy to produce from forecast/fable tools",
    "Good, but usually less production-oriented than Python",
    "Easy to create tables and plots for a report",
    "Use R as the main tool for this class project"
  ),
  Python = c(
    "Better for larger pipelines",
    "Automation, dashboards, ML, and production systems",
    "More setup and more manual structure",
    "Available, but takes more coding",
    "Strong for scheduled scripts and production workflows",
    "Useful, but needs more formatting work",
    "Use Python as a second tool when scaling the workflow"
  )
)
knitr::kable(r_python_comparison_final, caption = "R vs. Python forecasting comparison")
```

## Bonus #1 Recommendation

For Texas HealthLink, I would keep R as the main forecasting tool for analysis and reporting. R is faster for building a clear time series report with ETS, ARIMA, decomposition, diagnostics, and forecast plots. It is easier to audit and explain in a classroom or board report.

I would add Python as a second tool when the work needs to move beyond a report. Python is useful when HealthLink wants to automate the forecast, connect results to dashboards, combine forecasts with machine learning features, or deploy the process into a production system.

My recommendation is simple: use R for forecasting analysis and Python for scaling the workflow later.

## Bonus #1 Python Script Appendix

The Python script was run separately and produced the CSV outputs used in this section. The script is also submitted as a separate `.py` file.

```python
"""
DATA 624 Project 1 - Bonus #1
Texas HealthLink Python Forecasting Script - FIXED VERSION

Purpose:
This script supports Bonus #1 by reimplementing the Texas HealthLink
forecasting workflow in Python. It compares Python forecasting models,
selects the best model for each clinic using a holdout test, and produces
12-month forecasts for each clinic and the systemwide total.

Input needed:
healthlink_bonus1.csv

Required columns:
clinic, month, visits

Outputs created in a folder named bonus1_python_outputs:
1. python_bonus1_clinic_accuracy.csv
2. python_bonus1_clinic_forecasts.csv
3. python_bonus1_system_forecast.csv
4. python_bonus1_model_summary.csv
5. python_bonus1_report_summary.txt
6. python_bonus1_system_forecast_plot.png
7. python_bonus1_clinic_forecast_plot.png

Models used:
1. ETS / Exponential Smoothing
2. SARIMA / seasonal ARIMA-style model
3. Regression with trend and monthly seasonality

This script is designed to run outside RStudio. It does not need reticulate.

Author: Group 3
"""

from __future__ import annotations

import argparse
import warnings
from pathlib import Path

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

from sklearn.linear_model import LinearRegression
from sklearn.metrics import mean_absolute_error, mean_squared_error

from statsmodels.tsa.holtwinters import ExponentialSmoothing
from statsmodels.tsa.statespace.sarimax import SARIMAX


warnings.filterwarnings("ignore")


FORECAST_HORIZON = 12
HOLDOUT_MONTHS = 12


def rmse(y_true, y_pred):
    """Root mean squared error."""
    return float(np.sqrt(mean_squared_error(y_true, y_pred)))


def mape(y_true, y_pred):
    """Mean absolute percentage error."""
    y_true = np.asarray(y_true, dtype=float)
    y_pred = np.asarray(y_pred, dtype=float)
    mask = y_true != 0

    if mask.sum() == 0:
        return np.nan

    return float(np.mean(np.abs((y_true[mask] - y_pred[mask]) / y_true[mask])) * 100)


def make_month_features(dates):
    """Create monthly dummy variables."""
    dates = pd.to_datetime(dates)
    month_numbers = pd.Series(dates.month)
    month_dummies = pd.get_dummies(month_numbers, prefix="month", drop_first=True)
    return month_dummies.reset_index(drop=True)


def regression_forecast(train_series, forecast_dates):
    """
    Regression model with trend and monthly seasonal dummies.
    This is comparable to regression with seasonality in R.
    """
    train_dates = train_series.index

    train_x = pd.DataFrame({"trend": np.arange(1, len(train_series) + 1)})
    train_x = pd.concat(
        [train_x.reset_index(drop=True), make_month_features(train_dates)],
        axis=1
    )

    future_x = pd.DataFrame({
        "trend": np.arange(len(train_series) + 1, len(train_series) + len(forecast_dates) + 1)
    })
    future_x = pd.concat(
        [future_x.reset_index(drop=True), make_month_features(forecast_dates)],
        axis=1
    )

    future_x = future_x.reindex(columns=train_x.columns, fill_value=0)

    model = LinearRegression()
    model.fit(train_x, train_series.values)

    predictions = model.predict(future_x)
    return np.maximum(predictions, 0)


def ets_forecast(train_series, steps):
    """ETS / Exponential Smoothing forecast."""
    model = ExponentialSmoothing(
        train_series,
        trend="add",
        seasonal="add",
        seasonal_periods=12,
        initialization_method="estimated"
    )
    fit = model.fit(optimized=True)
    predictions = fit.forecast(steps)
    return np.maximum(np.asarray(predictions), 0)


def sarima_forecast(train_series, steps):
    """
    Seasonal ARIMA-style forecast using statsmodels SARIMAX.
    This uses a simple, stable model form for the class project.
    """
    model = SARIMAX(
        train_series,
        order=(1, 1, 1),
        seasonal_order=(1, 1, 1, 12),
        enforce_stationarity=False,
        enforce_invertibility=False
    )
    fit = model.fit(disp=False)
    predictions = fit.forecast(steps)
    return np.maximum(np.asarray(predictions), 0)


def clean_input_data(data_path):
    """Load and validate HealthLink data."""
    if not data_path.exists():
        raise FileNotFoundError(
            "\nCannot find the input data file.\n"
            f"Expected file: {data_path}\n"
            "Put healthlink_bonus1.csv in the same folder as this script, "
            "or pass the file path using --data.\n"
        )

    df = pd.read_csv(data_path)
    df.columns = [str(c).strip().lower() for c in df.columns]

    required_cols = {"clinic", "month", "visits"}
    missing_cols = required_cols - set(df.columns)

    if missing_cols:
        raise ValueError(
            "The data file is missing required columns: "
            + ", ".join(sorted(missing_cols))
        )

    df = df[["clinic", "month", "visits"]].copy()
    df["clinic"] = df["clinic"].astype(str).str.strip()
    df["month"] = pd.to_datetime(df["month"], errors="coerce")
    df["visits"] = pd.to_numeric(df["visits"], errors="coerce")

    bad_rows = df[df["month"].isna() | df["visits"].isna()]
    if len(bad_rows) > 0:
        raise ValueError(
            "Some rows have invalid month or visits values. "
            "Fix the CSV before running the script."
        )

    df = df.sort_values(["clinic", "month"]).reset_index(drop=True)

    return df


def get_monthly_series(df, clinic):
    """Return one monthly time series for one clinic."""
    clinic_df = df[df["clinic"] == clinic].copy()
    series = clinic_df.set_index("month")["visits"].sort_index().asfreq("MS")

    # Small safety rule: interpolate any missing month created by asfreq.
    # This should not happen if the data is complete, but it prevents a crash.
    if series.isna().any():
        series = series.interpolate(method="linear").ffill().bfill()

    return series


def evaluate_models_for_series(series, clinic_name):
    """Compare Python models using the last 12 months as the holdout set."""
    if len(series) <= HOLDOUT_MONTHS + 24:
        raise ValueError(f"Not enough data for holdout testing: {clinic_name}")

    train = series.iloc[:-HOLDOUT_MONTHS]
    test = series.iloc[-HOLDOUT_MONTHS:]
    forecast_dates = test.index

    model_predictions = {}

    model_functions = {
        "ETS": lambda: ets_forecast(train, HOLDOUT_MONTHS),
        "SARIMA": lambda: sarima_forecast(train, HOLDOUT_MONTHS),
        "Regression_Seasonality": lambda: regression_forecast(train, forecast_dates),
    }

    for model_name, model_func in model_functions.items():
        try:
            model_predictions[model_name] = model_func()
        except Exception as exc:
            print(f"{model_name} failed for {clinic_name}: {exc}")

    if not model_predictions:
        raise RuntimeError(f"All models failed for {clinic_name}.")

    rows = []
    for model_name, predictions in model_predictions.items():
        rows.append({
            "clinic": clinic_name,
            "model": model_name,
            "RMSE": rmse(test.values, predictions),
            "MAE": float(mean_absolute_error(test.values, predictions)),
            "MAPE": mape(test.values, predictions)
        })

    accuracy = pd.DataFrame(rows).sort_values(["RMSE", "MAE"]).reset_index(drop=True)
    best_model = accuracy.loc[0, "model"]

    return accuracy, best_model


def forecast_best_model(series, best_model, horizon):
    """Fit the selected model on all available data and forecast 12 months."""
    last_date = series.index.max()
    future_dates = pd.date_range(
        last_date + pd.offsets.MonthBegin(1),
        periods=horizon,
        freq="MS"
    )

    if best_model == "ETS":
        predictions = ets_forecast(series, horizon)
    elif best_model == "SARIMA":
        predictions = sarima_forecast(series, horizon)
    elif best_model == "Regression_Seasonality":
        predictions = regression_forecast(series, future_dates)
    else:
        raise ValueError(f"Unknown model: {best_model}")

    return future_dates, predictions


def create_plots(df, clinic_forecasts, system_forecast, output_dir):
    """Create system and clinic forecast plots."""
    historical_system = (
        df.groupby("month", as_index=False)["visits"]
        .sum()
        .rename(columns={"visits": "system_visits"})
    )

    plt.figure(figsize=(10, 5))
    plt.plot(
        historical_system["month"],
        historical_system["system_visits"],
        label="Historical system visits"
    )
    plt.plot(
        system_forecast["month"],
        system_forecast["system_total_forecast_visits"],
        label="Python forecast"
    )
    plt.title("Texas HealthLink Systemwide Visits: Python Forecast")
    plt.xlabel("Month")
    plt.ylabel("Visits")
    plt.legend()
    plt.tight_layout()
    plt.savefig(output_dir / "python_bonus1_system_forecast_plot.png", dpi=150)
    plt.close()

    plt.figure(figsize=(11, 6))
    for clinic in sorted(df["clinic"].unique()):
        actual = df[df["clinic"] == clinic]
        forecast = clinic_forecasts[clinic_forecasts["clinic"] == clinic]
        plt.plot(actual["month"], actual["visits"], alpha=0.65)
        plt.plot(forecast["month"], forecast["forecast_visits"], linestyle="--", alpha=0.9)

    plt.title("Texas HealthLink Clinic Forecasts: Python Models")
    plt.xlabel("Month")
    plt.ylabel("Visits")
    plt.tight_layout()
    plt.savefig(output_dir / "python_bonus1_clinic_forecast_plot.png", dpi=150)
    plt.close()


def write_report_summary(model_summary, system_forecast, output_dir):
    """Create a short text summary that can be pasted into the report."""
    best_models = (
        model_summary["selected_python_model"]
        .value_counts()
        .reset_index()
    )
    best_models.columns = ["model", "number_of_clinics"]

    avg_rmse = model_summary["holdout_RMSE"].mean()
    first_month = system_forecast["month"].min().strftime("%Y-%m")
    last_month = system_forecast["month"].max().strftime("%Y-%m")
    avg_system_forecast = system_forecast["system_total_forecast_visits"].mean()

    lines = []
    lines.append("Bonus #1 Python Forecasting Summary")
    lines.append("")
    lines.append("I reimplemented the Texas HealthLink forecasting workflow in Python.")
    lines.append("The Python workflow tested ETS, SARIMA, and regression with monthly seasonality.")
    lines.append("The last 12 months were used as a holdout period to compare model accuracy.")
    lines.append("")
    lines.append("Selected model count by clinic:")
    for _, row in best_models.iterrows():
        lines.append(f"- {row['model']}: {row['number_of_clinics']} clinic(s)")
    lines.append("")
    lines.append(f"Average clinic holdout RMSE: {avg_rmse:,.2f}")
    lines.append(f"Forecast horizon: {first_month} to {last_month}")
    lines.append(f"Average monthly systemwide forecast: {avg_system_forecast:,.0f} visits")
    lines.append("")
    lines.append("Main takeaway:")
    lines.append(
        "Python can reproduce the same forecasting workflow used in R. "
        "R is still stronger for direct time series reporting, while Python is useful "
        "when the forecast needs to connect with machine learning workflows, dashboards, "
        "or production systems."
    )

    report_text = "\n".join(lines)
    (output_dir / "python_bonus1_report_summary.txt").write_text(report_text, encoding="utf-8")


def main():
    parser = argparse.ArgumentParser(
        description="DATA 624 Bonus #1 Texas HealthLink Python Forecasting"
    )
    parser.add_argument(
        "--data",
        default="healthlink_bonus1.csv",
        help="Path to HealthLink CSV file. Default: healthlink_bonus1.csv"
    )
    parser.add_argument(
        "--output",
        default="bonus1_python_outputs",
        help="Output folder. Default: bonus1_python_outputs"
    )
    args = parser.parse_args()

    data_path = Path(args.data)
    output_dir = Path(args.output)
    output_dir.mkdir(parents=True, exist_ok=True)

    df = clean_input_data(data_path)

    print("\nDATA 624 Bonus #1 - Python Forecasting")
    print("--------------------------------------")
    print(f"Rows loaded: {len(df)}")
    print(f"Clinics: {', '.join(sorted(df['clinic'].unique()))}")
    print(f"Date range: {df['month'].min().date()} to {df['month'].max().date()}")
    print(f"Output folder: {output_dir.resolve()}")

    all_accuracy = []
    all_forecasts = []
    model_summary = []

    for clinic in sorted(df["clinic"].unique()):
        print(f"\nWorking on clinic: {clinic}")
        series = get_monthly_series(df, clinic)

        accuracy, best_model = evaluate_models_for_series(series, clinic)
        all_accuracy.append(accuracy)

        future_dates, predictions = forecast_best_model(series, best_model, FORECAST_HORIZON)

        forecast_df = pd.DataFrame({
            "clinic": clinic,
            "month": future_dates,
            "python_model": best_model,
            "forecast_visits": np.round(predictions, 0).astype(int)
        })
        all_forecasts.append(forecast_df)

        model_summary.append({
            "clinic": clinic,
            "selected_python_model": best_model,
            "holdout_RMSE": float(accuracy.loc[0, "RMSE"]),
            "holdout_MAE": float(accuracy.loc[0, "MAE"]),
            "holdout_MAPE": float(accuracy.loc[0, "MAPE"])
        })

    accuracy_out = pd.concat(all_accuracy, ignore_index=True)
    forecasts_out = pd.concat(all_forecasts, ignore_index=True)
    model_summary_out = pd.DataFrame(model_summary)

    system_forecast = (
        forecasts_out
        .groupby("month", as_index=False)["forecast_visits"]
        .sum()
        .rename(columns={"forecast_visits": "system_total_forecast_visits"})
    )
    system_forecast["method"] = "Bottom-up sum of clinic forecasts"

    accuracy_out.to_csv(output_dir / "python_bonus1_clinic_accuracy.csv", index=False)
    forecasts_out.to_csv(output_dir / "python_bonus1_clinic_forecasts.csv", index=False)
    system_forecast.to_csv(output_dir / "python_bonus1_system_forecast.csv", index=False)
    model_summary_out.to_csv(output_dir / "python_bonus1_model_summary.csv", index=False)

    create_plots(df, forecasts_out, system_forecast, output_dir)
    write_report_summary(model_summary_out, system_forecast, output_dir)

    print("\nFiles created:")
    for file_path in sorted(output_dir.iterdir()):
        print(f"- {file_path.name}")

    print("\nSelected Python models:")
    print(model_summary_out.to_string(index=False))

    print("\nDone. Use these outputs in Bonus #1 for the R vs Python comparison.")


if __name__ == "__main__":
    main()

```

# Bonus #2: Interview-Ready Time Series Literacy

## Executive Summary

When I start a forecasting problem, I do not jump straight into the model. I first look at the time series plot because the plot usually tells the first part of the story. I want to see whether the data is moving up or down, whether there is seasonality, and whether there are unusual periods that should not be ignored.

In this project, both datasets had a clear business purpose. Lone Star Appliances needed shipment forecasts to plan production, staffing, inventory, and logistics. Texas HealthLink needed patient visit forecasts to plan staffing, budgeting, scheduling, and clinic capacity.

The main lesson from both exercises is that forecasting is not just a technical task. The forecast has to connect back to a decision. For Lone Star, the 2020 drop showed why leadership should not rely only on the point forecast. For HealthLink, the clinic-level forecasts matter because each clinic has a different demand level. A systemwide forecast helps with budgeting, but the clinic forecasts are better for staffing.

My approach is simple. I check the pattern, test more than one model, compare the errors, review the residuals, and explain the forecast range in plain language. I do not treat the model as perfect. I treat it as a planning tool.

## Graph Commentary

The time series plots are the first check. They show whether the data is trending up, moving with a seasonal pattern, or showing unusual shocks. In Lone Star, the plot shows growth over time and stronger end-of-year demand. It also shows the 2020 disruption.

The decomposition plots help separate the series into trend, seasonal movement, and leftover noise. This is useful because leadership may only see one line moving up or down, but the model needs to separate the long-term direction from the repeating seasonal pattern.

The residual plots are the final check. I use them to see what the model missed. A good model should leave residuals that look mostly random. If the residuals still show a pattern, I would not be comfortable using that model without more review.

```{r bonus2-graph-1-timeseries}
autoplot(system_ts) +
  labs(
    title = "Interview Graph 1: Texas HealthLink Systemwide Visits",
    x = "Year",
    y = "Visits"
  )
```

In an interview, I would say this graph is the first thing I check. It shows the overall demand direction before I start defending a model.

```{r bonus2-graph-2-stl}
plot(system_decomp)
```

I would use this graph to explain the difference between trend, seasonality, and noise. This helps a nontechnical person understand why the model does not treat every month the same.

```{r bonus2-graph-3-acf-pacf}
Acf(system_ts, main = "Interview Graph 3A: System Total ACF")
Pacf(system_ts, main = "Interview Graph 3B: System Total PACF")
```

I would use these plots to explain autocorrelation in plain language. If past months are related to future months, the model should account for that pattern.

```{r bonus2-graph-4-residuals}
checkresiduals(hl_system_fit_for_diagnostics)
```

This is the graph I would use to explain model checking. A forecast is not finished just because the line looks reasonable. I still need to check what the model missed.

```{r bonus2-graph-5-forecast}
autoplot(hl_system_forecast) +
  labs(
    title = "Interview Graph 5: Systemwide Forecast with Uncertainty",
    x = "Year",
    y = "Visits"
  )
```

This is the graph I would show leadership. It gives the expected direction and the uncertainty range, which is more useful than a point forecast alone.

## Mock Interview Q&A

**Question: How do you start a forecasting problem?**  
I start with the plot. Before I choose a model, I want to understand the shape of the data. I look for trend, seasonality, outliers, and shocks. In this project, the Lone Star data showed long-term growth and a clear quarterly pattern. The HealthLink data showed clinic-level differences and systemwide growth.

**Question: What did you learn from the Lone Star forecast?**  
Lone Star shipments were increasing over time, but the 2020 period showed a disruption. That matters because a company can have strong demand and still face short-term shocks. I would use the forecast as a baseline, but I would also keep a buffer for inventory and logistics planning.

**Question: What did you learn from the HealthLink forecast?**  
The HealthLink forecast showed why clinic-level planning matters. The systemwide forecast is useful for the total budget, but staffing has to be planned by clinic. Temple and Waco have higher visit volume than smaller clinics, so one staffing rule would not fit every location.

**Question: Why compare ETS, ARIMA, regression, and seasonal naive models?**  
I do not want to trust one model too quickly. Seasonal naive gives a simple baseline. ETS is useful for trend and seasonality. ARIMA is useful when past values and autocorrelation matter. Regression with seasonality is easy to explain to leadership. Comparing them gives me a better reason for choosing the final model.

**Question: What does residual checking tell you?**  
Residuals show what the model missed. If residuals still have a pattern, that means the model may not be capturing the full story. In a business setting, I would not just show the forecast. I would also check whether the errors look reasonable.

**Question: How do you explain forecast uncertainty to leadership?**  
I explain that the point forecast is not a promise. It is the middle estimate. The interval is the planning range. If leadership only plans around the point forecast, they may be underprepared. For staffing, inventory, and logistics, the range matters.

**Question: What makes a forecast useful?**  
A forecast is useful when it helps someone make a better decision. In this project, the forecast supports production planning, clinic staffing, budgeting, inventory, logistics, and facility planning. The model is only one part of the work. The final step is explaining what leadership should do with it.

**Question: What could make the forecast wrong?**  
The forecast could be wrong if the future does not behave like the past. COVID-type disruptions, supply chain issues, staffing shortages, capacity limits, economic changes, and changes in customer or patient behavior could all affect the forecast.

**Question: Which tools would you use in practice?**  
For this project, R worked well because the forecasting package makes ETS, ARIMA, decomposition, and diagnostics straightforward. I would still use Python when the forecasting work needs to connect with larger machine learning pipelines, dashboards, or production systems. For a time-series class project, R is the cleaner tool.

# Final Conclusion

Both exercises point to the same main lesson: forecasting is useful only when it supports a decision. Lone Star should use the shipment forecast for production, staffing, inventory, and logistics planning. Texas HealthLink should use the clinic-level and systemwide forecasts for staffing, budgeting, scheduling, and capacity planning.

The models are not perfect, but they give leadership a better planning baseline than guessing. The safest approach is to review the forecast each period, compare the forecast to actual results, and update the plan when the data changes.

# Final Submission Checklist

This report includes the required project pieces:

- Exercise #1 board-level summary and technical appendix.
- Exercise #1 forecast models, validation, diagnostics, and recommendation.
- Exercise #2 board-level summary and technical appendix.
- Exercise #2 clinic-level forecasts and systemwide forecast.
- Exercise #2 forecast reconciliation using bottom-up totals.
- Bonus #1 R vs. Python forecasting comparison.
- Bonus #1 clean R and Python code appendix.
- Bonus #2 interview transcript, executive summary, and interview-ready graphs.
- Clear recommendations written for leadership.
- R code included in the report so it can be reviewed and rerun.

# References

CUNY School of Professional Studies. (2026). *DATA 624: Predictive Analytics syllabus* [Course syllabus].

CUNY School of Professional Studies. (2026). *Project #1: Lone Star Appliance* [Project instructions].

CUNY School of Professional Studies. (2026). *Data: Lone Star Appliance modeling data* [Dataset].

CUNY School of Professional Studies. (2026). *Project #1: Texas HealthLink* [Project instructions].

CUNY School of Professional Studies. (2026). *Data: HealthLink modeling data* [Dataset].

CUNY School of Professional Studies. (2026). *Bonus #1: Cross-Language Forecasting and Executive Comparison* [Bonus instructions].

CUNY School of Professional Studies. (2026). *Bonus #2: Interview-Ready Time Series Literacy* [Bonus instructions].

Hyndman, R. J., & Athanasopoulos, G. (2021). *Forecasting: Principles and Practice* (3rd ed.). OTexts.

Kuhn, M., & Johnson, K. (2013). *Applied predictive modeling*. Springer.

Statsmodels Developers. (n.d.). *Statsmodels time series analysis*. https://www.statsmodels.org/

Scikit-learn Developers. (n.d.). *Scikit-learn: Machine learning in Python*. https://scikit-learn.org/
